\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-11-03 paper.tex
Updates from Overleaf

% This is nice to have, but probably won't work in the final version
% without section numbers. Let's deal with that when we get there I guess.
We begin in Section~\ref{sec-gARG} by providing a precise formal definition
of a gARG, illustrated by an example ARG embedded in pedigree.
We then provide a similar definition of the classical eARG
approach in Section~\ref{sec-eARG}, and consider its limitations
in the context of current datasets and research questions.
Following this, we discuss the important concept of ancestral material
in Section~\ref{sec-ancestral-material}, and how it relates
to the process of converting an eARG to a gARG.
We continue
in Section~\ref{sec-ARG-and-local-trees} by considering the relationship
between an ARG and its local trees. Contrary to the
prevailing view, we show that a suitably encoded sequence of local trees contains
precisely as much information as the corresponding ARG.
The gARG encoding opens a rich new set of details about ARGs,
including the ideas of locally unary nodes
(Section~\ref{sec-locally-unary-edges}),
the levels of detail that can be represented in an ARG
(Section~\ref{sec-ARG-simplification}),
and the degrees of precision about recombination
that can be stored and we may seek to infer (Section~\ref{sec-precision}).
These ideas have important practical considerations,
which we illustrate by examining the qualitative properties of
ARGs inferred by four recent methods for a classical benchmark
dataset in Section~\ref{sec-example-inferred-args}.
We then discuss how the gARG framework can be
efficiently implementated in Section~\ref{sec-efficiency},
and finish with an assessment of the key challenges facing
the field in the Discussion.
Finally, the literature on ARGs is large and
confusing, and we attempt to clarify some important aspects
in appendices, including
a brief history of ancestral graphs (Appendix~\ref{sec-arg-history}),
a description of the Big and Little ARG stochastic processes
 (Appendix~\ref{sec-big-and-little-arg}),
a survey of ARG inference methods (Appendix~\ref{sec-survey-arg-infer}),
and a discussion of ARGs at an individual vs cell
lineage level (Appendix~\ref{sec-cell-lineages-and-args}).

@@

This combinatorial approach provides at least
one novel insight, clarifying the fundamental relationship between
ARGs and local trees.
Suppose we are given a gARG defined by a set of nodes and edges.
There is no requirement on the structure of this ARG beyond
the basic definitions: it could correspond to an ARG in which
every recombination event is exactly specified
(e.g.\ Fig.~\ref{fig-ancestry-resolution})
or one in which local trees are entirely disjoint
(i.e.\ only the sample nodes are shared between them).
If we  are given the sequence of local trees for this gARG
encoded as an oriented tree, along with the genome interval
covered by each tree, we can recover the original gARG exactly.
% We do this site-by-site for simplicity, but hopefully it's obvious
% that this isn't necessary?
More formally, suppose we are given the local tree $\pi^x_1\dots\pi^x_n$
for each nucleotide position $1 \leq x \leq L$ on a genome of length $L$.
Then, the edges of the ``local ARG'' for this tree is given by
$E^x = \{(u, \pi_u^x, \{x\}) \mid \pi_u^x \neq 0\}$. Because the ARG
edges are defined by $(c, p, I)$ tuples, where the set $I$ defines
the positions over which node $c$ inherits from parent $p$, we can
then simply combine the ``local ARGs'' for each position $x$
% YW: The above "combining" operation isn't obviously lossless to me:
% we could potentially lose information in the different
% combinations of intervals (e.g. if we use 2 non-combined intervals
% to represent a diamond in a 1-RE-node ARG). It seems like we might
% need an additional gARG spec to define what combinations are equivalent.
% But no-one else has commented on this, so I guess we just let it lie.
to recover precisely the same set of edges as the original ARG.
Thus, under this definition, there
is a one-to-one correspondence between an ARG and
the sequence of local trees that it encodes.

@@

% Here we describe recombination features
Considering Fig.~\ref{fig-inferred-args}, we can see that
there is substantial variation in the number of recombination breakpoints
inferred by different methods, with e.g.~\argweaver\ suggesting
far more than the 7 required for this dataset under minimal parsimony
assumptions~\citep{song2003parsimonious}.
A sense of the amount of recombination in each ARG is provided by
the node colouring scheme, which shows the number of parents for each node.
In Fig.~\ref{fig-inferred-args}A,B, each recombination event
corresponds to a node with exactly 2 parents and one child.
As these methods explicitly infer a recombination event for each
breakpoint, the number of breakpoints equals the number of 2-parent
(brown) nodes.
In contrast, Fig.~\ref{fig-inferred-args}C,D do not have this
straightforward
relationship between the number of nodes with multiple parents
and number of breakpoints along the genome.
In both ARGs the number of breakpoints is smaller
than the number of multiple-parent ARG nodes, showing
that several multiple-parent nodes must share breakpoint positions.
There are also ARG nodes with multiple parents and multiple children,
where one or more recombinations have been pushed down onto a
more recent node.
% Several of these nodes have more than two
% % parents (in red), representing multiple recombinations inferred to have
% occurred in an unresolvable order.
% Nevertheless, the number of breakpoints in
% these two ARGs is smaller than the number of multiple-parent ARG nodes, showing
% that several multiple-parent nodes must share breakpoint positions.
As a consequence, it may be difficult to condense each transition between trees
in these ARGs into a set of SPR operations.
% The sharing
% of breakpoint locations between these nodes indicates that in \tsinfer\ and
% \relate\ ARGs it may be difficult to condense each transition between trees
% into a set of simple SPR moves.
% Overall, Fig.~\ref{fig-inferred-args} shows
% that \tsinfer\ and \relate\ ARGs contain a level of detail about recombination
% that lies somewhere between a sequence of unrelated local trees on one extreme
% and an ARG with precisely specified recombination events on the other.

>>>>

% This is nice to have, but probably won't work in the final version
% without section numbers. Let's deal with that when we get there I guess.
We begin in Section~\ref{sec-gARG} by providing a precise formal definition
of a gARG, illustrated by an example ARG embedded in pedigree.
We then provide a similar definition of the classical eARG
approach in Section~\ref{sec-eARG}, and consider its limitations
in the context of current datasets and research questions.
Following this, we discuss the important concept of ancestral material
in Section~\ref{sec-ancestral-material}, and how it relates
to the process of converting an eARG to a gARG.
We continue
in Section~\ref{sec-ARG-and-local-trees} by considering the relationship
between an ARG and its local trees. Contrary to the
prevailing view, we show that a suitably encoded sequence of local trees contains
precisely as much information as the corresponding ARG.
The gARG encoding opens a rich new set of details about ARGs,
including the ideas of locally unary nodes
(Section~\ref{sec-locally-unary-edges}),
the levels of detail that can be represented in an ARG
(Section~\ref{sec-ARG-simplification}),
and the degrees of precision about recombination
that can be stored and we may seek to infer (Section~\ref{sec-precision}).
These ideas have important practical considerations,
which we illustrate by examining the qualitative properties of
ARGs inferred by four recent methods for a classical benchmark
dataset in Section~\ref{sec-example-inferred-args}.
We then discuss how the gARG framework can be
efficiently implemented in Section~\ref{sec-efficiency},
and finish with an assessment of the key challenges facing
the field in the Discussion.
Finally, the literature on ARGs is large and
confusing, and we attempt to clarify some important aspects
in appendices, including
a brief history of ancestral graphs (Appendix~\ref{sec-arg-history}),
a description of the Big and Little ARG stochastic processes
 (Appendix~\ref{sec-big-and-little-arg}),
a survey of ARG inference methods (Appendix~\ref{sec-survey-arg-infer}),
and a discussion of ARGs at an individual vs cell
lineage level (Appendix~\ref{sec-cell-lineages-and-args}).

@@

This combinatorial approach provides at least
one novel insight, clarifying the fundamental relationship between
ARGs and local trees.
Suppose we are given a gARG defined by a set of nodes and edges.
There is no requirement on the structure of this ARG beyond
the basic definitions: it could correspond to an ARG in which
every recombination event is exactly specified
(e.g.\ Fig.~\ref{fig-ancestry-resolution})
or one in which local trees are entirely disjoint
(i.e.\ only the sample nodes are shared between them).
If we  are given the sequence of local trees for this gARG
encoded as an oriented tree, along with the genome interval
covered by each tree, we can recover the original gARG exactly.
% We do this site-by-site for simplicity, but hopefully it's obvious
% that this isn't necessary?
More formally, suppose we are given the local tree $\pi^x_1\dots\pi^x_n$
for each nucleotide position $1 \leq x \leq L$ on a genome of length $L$.
Then, the edges of the ``local ARG'' for this tree is given by
$E^x = \{(u, \pi_u^x, \{x\}) \mid \pi_u^x \neq 0\}$. Because the ARG
edges are defined by $(c, p, I)$ tuples, where the set $I$ defines
the positions over which node $c$ inherits from parent $p$, we can
then simply combine the ``local ARGs'' for each position $x$
% YW: The above "combining" operation isn't obviously lossless to me:
% we could potentially lose information in the different combinations
% of intervals (e.g. if we use 2 non-combined intervals to represent a 
% diamond in a 1-RE-node ARG). It seems like we might need an additional
% gARG spec to define what combinations of intervals are equivalent.
% But no-one else has commented on this, so I guess we just let it lie.
to recover precisely the same set of edges as the original ARG.
Thus, under this definition, there
is a one-to-one correspondence between an ARG and
the sequence of local trees that it encodes.

@@

% Here we describe recombination features
Considering Fig.~\ref{fig-inferred-args}, we can see that
there is substantial variation in the number of recombination breakpoints
inferred by different methods, with e.g.~\argweaver\ suggesting
far more than the 7 required for this dataset under minimal parsimony
assumptions~\citep{song2003parsimonious}.
A sense of the amount of recombination in each ARG is provided by
the node colouring scheme, which shows the number of parents for each node.
In Fig.~\ref{fig-inferred-args}A,B, each recombination event
corresponds to a node with exactly 2 parents and one child.
As these methods explicitly infer a recombination event for each
breakpoint, the number of breakpoints equals the number of 2-parent
(brown) nodes.
In contrast, Fig.~\ref{fig-inferred-args}C,D do not have this
straightforward
relationship between the number of nodes with multiple parents
and number of breakpoints along the genome.
In both ARGs the number of breakpoints is smaller
than the number of multiple-parent ARG nodes, showing
that several multiple-parent nodes must share breakpoint positions.
There are also ARG nodes with multiple parents and multiple children,
where one or more recombinations have been pushed down onto a
more recent node.
% Several of these nodes have more than two
% % parents (in red), representing multiple recombinations inferred to have
% occurred in an unresolvable order.
% Nevertheless, the number of breakpoints in
% these two ARGs is smaller than the number of multiple-parent ARG nodes, showing
% that several multiple-parent nodes must share breakpoint positions.
As a consequence, it may be difficult to condense each transition between trees
in these ARGs into a set of SPR operations.
% The sharing
% of breakpoint locations between these nodes indicates that in \tsinfer\ and
% \relate\ ARGs it may be difficult to condense each transition between trees
% into a set of simple SPR moves.
